---
title: "Final Project"
author: "Dexter Schincke, Emmanuel Rayappa"
date: "12/10/2021"
output: pdf_document
---

# Introduction

We have the challenge of convincing first time fliers that planes are safe. These people have never taken a trip on a plane and are nervous.Talking about aviation safety is important because there are certain people who tend to take cars more often than planes because they think that planes are less safe than cars, when in reality this is the opposite. I also had a brief fear of flying until I took my first Solo-Trip to San Jose and later on discovered that planes are much safer than people realize.

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(hablar)
```

# Main Focus and method
As there are many types of aircraft out there (Military, Private, Cargo, Passenger, etc), we want to narrow down our focus to the two most popular companies, which are Boeing (based in the United States), and Airbus(based in France.) 

To be able to analyze all the accidents that have happened involving theses two companies, we will be using a data set from Kaggle.com, which documents every accident that has occurred from 1908 to 2019. As mentioned earlier, we will only be focusing on accidents involving planes that are built by either Boeing of Airbus. This will require many data visualizations comparing the various number of accidents, number of fatalities (seeing if location was a factor), analyzing the ratio between the number of people on board the aircraft against the number of people who reported to have injuries. We will be using side by side box plots to assess the fatalities of the different types of aircraft, scatter plots to see how the ratios of the different number of fatalities compare, bar plots to count the number of reported incidences based on location, and time frame to see if there is a correlation, and a few additional graphs to help utilize the data frame. We will use key variables such as aircraft type, operator, location, date, and many more in our analysis.

```{r, eval = FALSE}
accidents <- read.csv("Airplane_Crashes_and_Fatalities_Since_1908_20190820105639.csv")

accidents[accidents == "NULL"] = 0

accidents_revised = accidents

cols.num <- c("Aboard","Aboard.Crew", "Aboard.Passangers", "Fatalities.Passangers", "Fatalities.Crew", "Ground", "Fatalities")

accidents_revised[cols.num] <- sapply(accidents_revised[cols.num],as.numeric)

glimpse(accidents_revised)

accidents_clean <- accidents_revised %>% select(Date, Time, Location, Operator, AC.Type, Aboard, Aboard.Passangers, Aboard.Crew, Fatalities, Fatalities.Crew, Fatalities.Passangers, Ground)

accidents_clean <- accidents_clean %>% mutate(Fatalities.Total = Fatalities + Ground)

accidents_BA_ABU <- accidents_clean %>% filter(str_detect(AC.Type, 'Boeing|Airbus'))

glimpse(accidents_BA_ABU)

view(accidents_BA_ABU)
```


Here we have filtered out all the accident data involving them. We can first start by looking at the number of accidents for Boeing aircraft and group it by the model. 

```{r}
accidents_Boeing <- accidents_clean %>% filter(str_detect(AC.Type, 'Boeing'))
accidents_Airbus <- accidents_clean %>% filter(str_detect(AC.Type, 'Airbus'))

glimpse(accidents_Boeing)
```

In order to properly represent the data, we decided to look at the Operators for the Boeing data that had 10 or more entries. This ensures that the data for Boeing is going to be more precise and accurate as it only looks to the more frequent Operators with reported hull losses and the data for the Boeing accidents was 10 times larger than for Airbus accidents.

```{r}
frequent_operatorsB <- accidents_Boeing %>% group_by(Operator) %>%
    filter(n() >= 10)
```

```{r}
ggplot(frequent_operatorsB, aes(x=Operator)) + geom_bar()
ggplot(frequent_operatorsB, aes(x = Fatalities.Crew)) + geom_boxplot(aes(fill = Operator)) + coord_flip()
ggplot(frequent_operatorsB, aes(x = Fatalities.Passangers)) + geom_boxplot(aes(fill = Operator)) + coord_flip()
```

From the plot about top operators, we can see that the top operators of Boeing Aircraft are American Airlines, United Airlines, and Pan American World Airways. The United States Army and Air Force are also top users of Boeing Aircraft. 

For Passenger Fatalities, American Airlines had the highest average number of fatalities for passengers at 50 while Pan American had 25 and United Airlines had 10. The United States Army and Air Force had almost zero because in most of their accidents, the people on board are classified as "Crew" rather than as "passengers". 

For Crew Fatalities, The United States Air Force has the highest number of fatalities among the Crew followed by American Airlines. American Airlines had the second highest average followed by Pan American and United, who were tied for the third highest followed by the US Army with the lowest average. 

Before doing the analysis, I thought that the carrier's that had the highest average fatalities among passengers would also have the highest among crew, but that wasn't the case as the US Military have Crew rather than passengers. 

We can repeat the process that we followed to analyze fatalities by passengers and crew for Airbus Aircraft. It is worth noting that their is a higher chance that there will be more passenger Airlines as Airbus isn't know for their military aircraft, but rather their passenger aircraft. 
```{r}
frequent_operatorsA <- accidents_Airbus %>% group_by(Operator) %>%
    filter(n() >= 2)
```

```{r}
ggplot(frequent_operatorsA, aes(x=Operator)) + geom_bar()
ggplot(frequent_operatorsA, aes(x = Fatalities.Passangers)) + geom_boxplot(aes(fill=Operator)) + coord_flip()
ggplot(frequent_operatorsA, aes(x = Fatalities.Crew)) + geom_boxplot(aes(fill = Operator)) + coord_flip()
```
We filter out the Airbus Data with the most frequent operators and find that Air France is a frequent operators of Airbus Aircraft.

From the passenger data, we see that China Airlines has the highest average of fatalities among passengers While American Airlines has the second highest, followed by Thai Airways, Indian Airlines, Egypt Air, Air France, and Philippine Airlines. It is  worth noting that Air France was the only carrier to have an out-lier, which is due to the crash of Air France Flight 447 in 2009 that killed 216 Passenger and 12 Crew after crashing into the Atlantic Ocean 700 miles off the coast of Brazil.

From the crew data, we see that China Airlines also had the highest average followed by Thai Airways,Egypt Air,American Airlines,Indian Airlines, Air France and Philippine Air Lines. 

Now there are also accidents in which Ground Fatalities occur and none of the previous analysis prove that there is an Airlines with high fatalities (Passengers, Crew, or Ground). So what we will do now is use data based on the Fatalities variable, which combines Passengers, Crew, and Ground fatalities. 
```{r}
ggplot(frequent_operatorsA, aes(x = Fatalities)) + geom_boxplot(aes(fill = Operator)) + coord_flip()
ggplot(frequent_operatorsB, aes(x = Fatalities)) + geom_boxplot(aes(fill = Operator)) + coord_flip()

```
From the plots, we see that American Airlines has the highest number of fatalities amongst passengers and crew. All in all, people who do not enjoy flying will avoid American Airlines as they have proven to have a number of accidents which causes them to have a high mean for fatalities. 


```{r, message=FALSE, warning=FALSE}
ggplot(frequent_operatorsA, aes(x = Fatalities.Total, y = Aboard)) + geom_point(aes(color=Operator)) + geom_abline(slope = 1)
ggplot(frequent_operatorsB, aes(x = Fatalities.Total, y = Aboard)) + geom_point(aes(color=Operator)) + geom_abline(slope = 1) + xlim(0,400)
```

One important thing to denote by the data visualization for the Aircraft Type Airbus is that we are able to see how linear the passenger aboard is compared to the total Fatalities. Here we are able to see that a larger proportion of people that were considered to be fatalities during the crash. Looking at the data on the left most side of the graph, we are able to conclude that there is reported to be a large number of people aboard the plane and there are not a lot of reported fatalities. This is good! The more data entries that are on the left most side of the graph is considered to be a much better ratio than the linear line. The linear line is at a slope of 1, which means that there is a ratio of 1:1 for people on the plane and number of fatalities.

Now we look at the proportions of people aboard the plane for the Aircraft Type Boeing. Looking at this data, we are able to see that there is a much larger number of planes that fall on the linear line. This means that there is a large portion of planes that are having a 1:1 ratio for people on board the plane and the number of fatalities. This shows that the U.S. Air Force and Army are the two main operators contributing to the ratio of the 1:1 where people on board the plane were often injured or killed when the plane crashed.

The data seems to be very contradicting. The Airbus data makes flying seem safe while the Boeing data seems to show that if you were involved in a Boeing plane accident, you would most likely be a fatality statistic. This makes the information seem a bit worrying. However, if we dive more into the Boeing ratios of people that were injured for the crew and for the passengers, we are able to sort through the confusion.

```{r}
ggplot(frequent_operatorsB, aes(x = Fatalities.Passangers, y = Fatalities.Crew)) + geom_point(aes(color=Operator))
```

Now we are able to recognize the importance of the ratios. In the previous graph, we noticed that the Boeing Aircraft type for the U.S. Air Force and Army made up a large portion of the 1:1 ratios. This information shows that a large majority of those planes had only military personnel and crew aboard those planes. If only crew were aboard the plane during the planes' crashes, the crew aboard the plane would count for the fatalities. This now makes sense why the previous data visualization had Military crashes as a 1:1 ratio for people aboard and fatalities. 

One big improvement we could make in this experiment is that we need data that is even in size as the sample for Boeing data was 10 times larger than for Airbus. We also found that some accidents did not appear on this data set, which could affect the results.
