---
title: "Lab 4: Visualizing Weather Data"
subtitle: "MTH 365: Intro to Data Science"
author: "Your Name Here"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

When we first introduced the `weather` dataset in class, many of you had some... concerns... about how the data was formatted and presented.

I didn't disagree.

For today's lab, you'll use what you've learned about data wrangling to modify this data set. You'll also explore some customization options for plots in `ggplot2`.

```{r}
library(tidyverse)
library(mosaicData)
data(Weather)
glimpse(Weather)

# This line replaces the NAs for weather events with the string 'None'
Weather <- Weather %>% mutate(events=ifelse(is.na(events), 'None', events))
```

## Dealing with weather events

> 1. How many distinct weather events were recorded in this data (ex. "Rain" and "Snow" are distinct events)? How many weather event combinations are possible?

```{r}
 Weather %>%
group_by(events) %>%
summarise(N =n())
```

`dplyr` comes with a set of helper functions that can help you select groups of variables inside a `select()` command:

- `starts_with("X")`: every name that starts with "X"
- `ends_with("X")`: every name that ends with "X"
- `contains("X")`: every name that contains "X"
- `matches("X")`: every name that matches "X", where "X" can be a regular expression
- `num_range("x", 1:5)`: the variables named x01, x02, x03, x04 and x05
- `one_of(x)`: every name that appears in x, which should be a character vector

Another useful function is the `str_detect` function: this returns a value of `TRUE` if a _string_ of characters is contained within another string.

```{r}
string <- 'abcdefg'
str_detect(string, 'abc')
str_detect(string, 'h')
```

> 2. Create a new variable for each weather event that records whether or not a specific event occurred on each day. For example, a `Rain` variable might take possible values of `TRUE` and `FALSE`.

```{r}
Weather = Weather %>%
mutate(Rain = str_detect(events,"Rain"))
glimpse(Weather)

Weather = Weather %>%
  mutate(Snow = str_detect(events, "Snow"))
  glimpse(Weather)

Weather = Weather %>%
  mutate(Fog = str_detect(events, "Fog"))
  glimpse(Weather)
  
Weather = Weather %>%
  mutate(Thunderstorm = str_detect(events, "Thunderstorm"))
  glimpse(Weather)

Weather = Weather %>%
mutate(Hail = str_detect(events, "Hail"))
glimpse(Weather)

```

> 3. Create a new variable that records the _number_ of weather events that occurred on each day.

```{r}
Weather_rain = Weather %>%
  mutate(count = Thunderstorm +Rain + Snow + Hail + Fog)
glimpse(Weather_rain)

```

> 4. Which weather events occurred most and least frequently? Does the answer depend on city and or year?

Rain had the most occurrences. Snow had the 2nd most occurrence followed by fog and hail. The answer does depend on the location as the city of Mumbai had 0 snow days and Mubai isn't in a favorable location for snow. 

```{r}
Weather %>%
  group_by(city, year) %>%
  summarise(Nrain = sum(Rain), Nsnow = sum(Snow), NFog = sum(Fog), NThunderstrom = sum(Thunderstorm), NHail = sum(Hail))
  


  

```

> 5. Calculate the average number of weather events per day for each city.

```{r}
Weather %>%
  group_by(city, day) %>%
  summarise(Ncount = mean(count))

```

> 6. Where and when did the days with the most weather events occur? What events were they? Make a table with city, date, and the number of weather events.

The days with the most weather events (3 was the max) occured in 2016 and 2017 in all the cities except Mumbai.
```{r}
Weather_rain %>%
  select(city, date, count, events) %>%
  arrange(desc(count))
  

```

## Explore color and theme options

There are lots of available color libraries in R that interact nicely with `ggplot2`. Today we will explore three: `RColorBrewer`, `viridis`, and `ggsci`.

### `RColorBrewer`

`RColorBrewer` takes the classic "color brewer" palettes (http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) and imports them into R. The color brewer palettes are part of most graphing tools. 

- https://www.r-graph-gallery.com/38-rcolorbrewers-palettes/

> 7. The added layer commands `scale_fill_brewer()` and `scale_color_brewer()` can be used select a color palette from the `RColorBrewer` library. Modify the plots below to use two new color palettes of your choice.

```{r}
#install.packages('RColorBrewer')
library(RColorBrewer)

Weather %>% 
ggplot(aes(y=avg_temp, x=city)) +
geom_boxplot(aes(fill=city)) + 
scale_fill_brewer(palette='Set3')

Weather %>% ggplot(aes(x=low_temp, y=high_temp)) + geom_point(aes(col=city)) + scale_colour_brewer(palette='Set2')
```

### `viridis`

Another option is the color palettes in the `viridis` library. Use the color scales in this package to make plots that are pretty, better represent your data, easier to read by those with colorblindness, and print well in grey scale.

There are only four color scales in the `viridis` package: `viridis`, `magma`, `plasma`, and `inferno`. 

> 8. The added layer commands `scale_fill_viridis()` and `scale_color_viridis()` can be used select a color palette from the `viridis` library. Modify the plots below to use two new color palettes of your choice.

```{r}
#install.packages('viridis')
library(viridis)

Weather %>% ggplot(aes(y=avg_temp, x=city)) +
  geom_boxplot(aes(fill=city)) + 
  scale_fill_viridis(option='magma', discrete = TRUE)

Weather %>% ggplot(aes(x=low_temp, y=high_temp)) + geom_point(aes(col=city)) + scale_colour_viridis(option='plasma', discrete=TRUE)
```

> 9. What error do you get if you remove the `discrete=TRUE` option? What do you think it means?

Discrete is used to define the type of data. By removing it, you will be told that you are trying to use a categorical data set.  


### `ggsci`

A third option is the `ggsci` package. This package contains color palettes from major scientific journals and sci-fi TV shows. A good mix.

- https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html#npg

> 10. `ggsci` modifies the colors using `scale_fill_XXX` and `scale_color_XXX`. Modify the plots below to use two new color palettes of your choice.

```{r}
#install.packages('ggsci')
library(ggsci)
library(ggplot2)
library(gridExtra)
data("Weather")

p1 = ggplot(Weather,aes(x=avg_temp)) +
  geom_histogram(aes(fill=city)) + 
  theme_bw()

p2 = ggplot(Weather,aes(x=low_temp, y=high_temp)) + geom_point(aes(col=city)) + theme_bw()

p1_npg = p1 + scale_color_npg()
p2_npg = p2 + scale_color_npg()
grid.arrange(p1_npg, p2_npg, ncol = 2)

```

### Themes

You can customize the plotting window in `ggplot2` using one of the built-in themes or by building your own.

> 11. Copy the plots with your favorite color schemes from question 7 to 10, and change the theme to two of your choosing from the list below. Which looks "best" with your preferred color scheme, and why?

- `theme_gray()` *default
- `theme_bw()`
- `theme_linedraw()`
- `theme_light()`
- `theme_dark()`
- `theme_minimal()`
- `theme_classic()`
- `theme_void()`

```{r}
Weather %>% ggplot(aes(y=avg_temp, x=city)) +
  geom_boxplot(aes(fill=city)) +
  theme_dark()

Weather %>% ggplot(aes(x=low_temp, y=high_temp)) + geom_point(aes(col=city)) +
theme_minimal()

```

> 12. The code below modifies the theme manually. Make two plots that are as ugly as you can by modifying the themes. 

```{r}
Weather %>% ggplot(aes(y=avg_temp, x=city)) +
  geom_boxplot(aes(fill=city)) +
  theme(panel.background = element_rect(fill = "pink",
                                colour = "pink",
                                size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                colour = "white"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                colour = "white")
  )

Weather %>% ggplot(aes(x=low_temp, y=high_temp)) + 
  geom_point(aes(col=city)) +
  theme(panel.background = element_rect(fill = "green",
                                colour = "green",
                                size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                colour = "red"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                colour = "blue")
  )
```

-------

Question 1-6 worth 10 points. Question 7-12 worth 5 points

Your submissions on question 1-6 will be assessed on:

1. Correctness of your code and whether has addressed the question. (6 pts. per question).
3. Code formatting: consistency, readability, etc. (4 pts. per question).

Question 7-12 will be assessed on completeness. 

Total points possible: 90