---
title: 'Lab 8: Building a Spam Filter'
author: "Emmanuel Rayappa"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document: default
  html_document: default
subtitle: 'MTH 365: Intro to Data Science'
---

So far we've seen three techniques for classification:

1. Classification trees and random forests
2. $k$-nearest neighbors
3. Artificial neural networks

Let's try all three on a single task, and see which one produces the "best" results.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(mosaic)

# install.packages('openintro')
library(openintro)
data(email)

email <- email %>% select(-time) %>% na.omit()
glimpse(email)
```

> 1. Create a testing and training data set, using an 80-20 split (80% training, 20% testing).

```{r}
test_id <- sample(1:nrow(email), size=round(0.20*nrow(email)))
TEST <- email[test_id,]
TRAIN <- email[-test_id,]

```


## Technique 1: Random forests and classification trees

> 2. Use a random forest to determine which variable(s) are most important for classifying an email as spam.

The variables that are most important for classifying an email as "spam" are num_char and line_breaks. The error rate for the forest is just 6.58%
```{r}
library(randomForest)
set.seed(365)
forest <- randomForest(as.factor(spam)~.,data=email, 
                       ntree=201, mtry=3)

forest

varImpPlot(forest)
```


> 3. Build a classification tree using these important variables.

There is a clear relationship between the number of characters and line breaks for emails to be considered as spam. 
```{r}
library(rpart)
library(rattle)
library(rpart.plot)
tree <- rpart(spam~num_char+line_breaks,data = TRAIN)
fancyRpartPlot(tree)
```


> 4. Apply the classification tree to your testing data set, and build a confusion matrix. How well does the classification tree categorize emails into spam or non-spam?

```{r}
test_x <- TEST %>% select(num_char,line_breaks)
TEST <- TEST %>% 
  mutate(tree_spam = predict(tree,test_x,type="class"))
confusion_test <- mosaic::tally(spam~tree_spam, data = TEST)

sum(diag(confusion_test))/nrow(TEST)

```

## Technique 2: $k$-nearest neighbors

> 5. $k$-nearest neighbors assumes that all explanatory variables are numeric. Before fitting the model, you'll need to remove two variables from your training and testing data. Which ones?

The two variables that we will need to remove are "winner" and "number".

```{r}
email_2 <- email %>% select(-winner) %>% na.omit() %>% select(-number) %>% na.omit()
test_id <- sample(1:nrow(email_2), size=round(0.20*nrow(email_2)))
TEST2 <- email_2[test_id,]
TRAIN2 <- email_2[-test_id,]

```


> 6. Use $k$-nearest neighbors on the training data set to classify emails as spam or non-spam. 

```{r}
library(class)
knn10<- knn(TRAIN2[,-1], test= TEST2[,-1], 
            cl=TRAIN2$spam, k=11, prob=TRUE)
glimpse(knn10)

```

> 7. Apply the $k$-nearest neighbors technique, and build a confusion matrix. How well does the $k$-nearest neighbors algorithm categorize emails into spam or non-spam?

```{r}
train_x <- TRAIN %>% select(num_char,line_breaks)
predicition_knn <- knn(train_x, test = test_x, cl =TRAIN$spam, k = 11, prob = TRUE)
confusion_test_2 <- mosaic::tally(spam~predicition_knn, data = TEST)

sum(diag(confusion_test_2))/nrow(TEST)




```

## Technique 3: Artificial neural networks

> 8. Build a single-layer artificial neural network with 20 nodes. 

```{r}
library(nnet)
spam_nnet <- nnet(spam~line_breaks + num_char, 
                    data=TRAIN, size= 20)
spam_nnet
```


> 9. Apply the ANN to your testing data, and build a confusion matrix. How well does the ANN categorize emails into spam or non-spam?

```{r}
TEST <- TEST %>% 
  mutate(nnet_spam = predict(spam_nnet,test_x,type="class"))
confusion_test_3 <- mosaic::tally(spam~nnet_spam, data = TEST)

sum(diag(confusion_test_3))/nrow(TEST)

```

## Recommendation

> 10. Of the three techniques you tested, which do you recommend for implementing a spam filter? Explain your recommendation.

The Knn test is the most accurate as it has a 90% accuray rate, the highest of the 3 techniques tried.