---
title: 'Mini-Project 1: Comic characters analysis'
author: "Your Name Here"
date: "Due September 21, 2021"
output:
  pdf_document: default
  html_document: default
subtitle: 'MTH 365: Intro to Data Science'
---

In this project you'll use data visualization tools to explore comic characters through the data comes from Marvel Wikia and DC Wikia. The two data (Marvel and DC are seperated) are available on BlueLine as `marvel-wikia-data.csv` and `dc-wikia-data.csv`. Please directly download the data from BlueLine instead of from the website. 

- Source: W. Hickey. "Comic Books Are Still Made By Men, For Men And About Men". Data description available on [Comic Characters](https://datahub.io/five-thirty-eight/comic-characters#readme)

> 0. Load the necessary packages and both datasets into RStudio. How many characters are there in each data? To simplify the visualization, we only keep the characters will full information. Write a paragraph to describe the project. 

```{r}
library(tidyverse)

marvel = read.csv("marvel-wikia-data.csv", na.strings=c("","NA"))
dc = read.csv("dc-wikia-data.csv", na.strings=c("","NA"))

marvel = marvel[complete.cases(marvel),]
dc = dc[complete.cases(dc),]
```

Use the data to answer the following research questions. For each question, you should produce _at least one data visualization_, with appropriate axis and legend labels (if necessary). Each data visualization should be accompanied by a paragraph describing the insights to be gained. 

-------

> 1. For marval comic, which years we have most characters created and which years we have fewest? Comment on the distribution of the first appreance year. 

With the marvel comics, we can see that they are most frequent after the year 2000.

```{r, fig.cap = "Density plot showing the trend of characters made every year"}
ggplot(marvel, aes(x = YEAR)) + geom_density() + 
xlab("Year") + 
ylab("Frequency")

```

> 2. We define the common characters are the ones with appearances more than 100 times. What is the median value of the appearance for the common characters in the marvel comic? Is there anyone appear much more often than the others? 

The average number of apperance according to this box plot is approximately 200 apperances. There are several outliers in thsi dataset which deos suggest that there are some marvel characters who make a lot more apperances than others.

```{r, fig.cap = "Box plot showing the trend of character apperance over 100"}
filter_apperances <- marvel %>% filter(APPEARANCES > 100)
ggplot(filter_apperances, aes(x = APPEARANCES)) + 
geom_boxplot() + 
xlab("Number of apperances") + 
ylab("Frequency")
```

> 3. What is the most common type of align among all marvel character? What is the most common type of align among the common character? 

The most common type of align amongst all marvel characters is good characters.

```{r, fig.cap = "Bar plot showing the various aligns and their frequencies"}
ggplot(marvel, aes(y = ALIGN)) + 
geom_bar(aes(fill = ALIGN)) + 
xlab("Frequency") + 
ylab("Character Type") 
```

> 4. How are the number of appearances and year being created are related in marvel comic? Calculated the correlation coefficient. Is there a linear relationship?

There isn't a linear relationship.
```{r, fig.cap = "Scatter plot showing the trend of characters made every year"}
ggplot(filter_apperances, aes(x = YEAR, y = APPEARANCES)) + 
geom_point()+
xlab("Year")+
ylab("Number of Apperances")

```

> 5. Add another variable to the marvel data which indicates whether the character is created after 1980. Compare the proportion of the gender distribution of characters before and after 1980 for marvel comic. 
(Hint: in the aesthetics, try `y=..prop..`)

In the data before 1980, the number of male characters to female characters had quite a large difference. In the data after 1980, the trend is the exact same.
```{r, fig.cap = "Bar graph for gender trends before and after 1980"}
marvel_1980 <- marvel %>%
  mutate(After =  YEAR > 1980)
ggplot(marvel_1980, aes(x = SEX)) + geom_bar(aes(group = SEX), bins = 30)+ xlab("Sex") + ylab("Count")

marvel_1979 <- marvel %>%
  mutate(Before = YEAR < 1980)
  ggplot(marvel_1979, aes(x = SEX)) + geom_bar(aes(group = SEX), bins = 30) + xlab ("Sex") + ylab("Count")
```


> 6. Among DC characters who appears more than 100 times and the eye colors are either black, brown, blue or green, show the distribution of appearances by each eye color. The plot color should matches the eye color. Comment the difference. 

All four have a decline of some sort with black eyes having the steepest decline while brown, green, and blue have a less steep decline.

```{r, fig.cap = "Comapring Eye color trends for charcters with more than 100 apperances"}

filter_apperance_dc <- dc %>% filter(EYE %in% c("Blue Eyes", "Brown Eyes", "Black Eyes", "Green Eyes"), APPEARANCES > 100)

ggplot(filter_apperance_dc, aes(x = APPEARANCES)) + geom_density(aes(fill = EYE)) + facet_wrap(~EYE, nrow = 2) +
  scale_color_manual(breaks = c("Blue" , "Brown", "Black", "Green"),
  values = c("blue", "brown", "black", "green")) +
  ylab("Frequency")


```

> 7. Combine marvel data and dc data. Compare the distribution of character identities in two comics.  (Hint: You can only combine the data when they have exactly the same column names. In the new data, you need an additional variable to indicate whether the data is Marvel or DC)

The distributions for DC only feature 3 different kinds of identities(Unknown, Public, or secret) while the marvel identities features all 7 types. Generally, there is no consistent trend for each identity as they vary in either a positive or negative skew.
```{r, fig.cap="Density plot comparing IDs for Marvel and DC Characters"}
dc <- dc %>% mutate(universe = "DC")
marvel <- marvel %>% mutate(universe = "Marvel")
marvel_and_dc <- bind_rows(dc,marvel)
ggplot(marvel_and_dc, aes(x = ID)) + geom_density(aes(color= ID))+facet_wrap(~universe, nrow = 2) + xlab("ID Type") + ylab("Frequency")

```

-------

Question 0 worth 8 points and will be assessed on the quality of written responses. Question 1-4 worth 8 points each and question 5-7 worth 20 points each. The questions will be assessed on 

1. Quality of the data visualizations (50\%).
2. Quality of your written responses (25\%).
3. Code formatting: consistency, readability, etc. (25\%).

Total points possible: 100