---
title: "Lab 10"
author: "Emmanuel Rayappa"
date: "Updated `r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    theme: cosmo
    code_download: yes
subtitle: 'MTH 362: Statistical Modeling'
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

Answer the questions below (use blockquotes to denote your answer by putting a ">" at the beginning of the line). When you're finished:

1.  Change your name in the "author:" space at the top of this document.
2.  Click the "Knit" button at the top to create an PDF/HTML file with your answers.
3.  Review your answers, make any changes as necessary, and re-"Knit".
4.  Save your PDF/HTML file and upload to BlueLine.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(MASS)
library(faraway)
library(lme4)
```

## Question 1

The `ratdrink` data set consists of five weekly measurements of body weight for 27 rats. The first 10 rats are on a control treatment, while 7 rats have thyroxine added to their drinking water. 10 rats have thiouracil added to their water.

```{r}
data(ratdrink)
head(ratdrink)
```

a) Make an appropriate plot of the data. Your plot should include age, treatment, and weight. What can you learn from your plot?
```{r}
ratdrink %>% ggplot(aes(x = weeks, y = wt)) + geom_point(aes(col = treat))
```

b) Fit a linear repeated measures model that allows for a random slope and intercept for each rat. Each group should have a different mean line.
```{r}
model1 = lmer(wt~weeks + treat + (weeks|subject), data = ratdrink)
summary(model1)
```

c) Interpret the fixed effect intercept term of your model.
> The fixed effect intercept is when the treatment is control and the week is at 0, the rat's expected weight is 52.98

d) Interpret the interaction between thiouracil and week.
> Compared to the control group, the weekly increased weight for thiouracil is 9.37 less

e) Interpret the random effect standard deviation.
>The subject will randomly affect the weight by about 5.74 units. It will also affect the relation between weeks and weight by approximately 3.8 units. 

f) Check whether there is a significant treatment effect.
> Since the p value is small, there is a significant treatment effect.

```{r}
model_alt = lmer(wt~weeks * treat + (weeks|subject), data = ratdrink, REML = FALSE)
model_null = lmer(wt~weeks + (weeks|subject), data = ratdrink, REML = FALSE)
library(pbkrtest)
KRmodcomp(model_alt, model_null)
```


g) Make diagnostic plots showing the residuals against the fitted values and a Q-Q plot of the residuals. Interpret your plots. 
> For the residual plots, the points are equally distributed above and below the zero line and there is no clear change on the variance. For the QQ plot, all the points are close enough to the 45 degree line, thus the linear assumptions are all met. 

```{r}
plot(model1)
library(lattice)
qqmath(model1)
```

h) Construct confidence intervals for the model parameters. Which random effect terms may not be significant?
> The interaction between random intercept and random slope may not be significant. 

```{r}
confint(model1, oldNames = FALSE)
```

i) Are either of the treatment groups "significantly different" from the control group?
> The thiouracil group is significantly different from the control group on the effects of weekly increased weight. 

## Question 2

The `attenu` data set gives peak acceleration measured at various observation stations for 23 earthquakes in California. The data has been used by various research teams to estimate the attenuating effect of distance on ground acceleration.

```{r}
data(attenu)
head(attenu)
```

a) Make a plot showing how the acceleration increases with distance for each quake. Use appropriate transformations of both axes so that the relationship is easier to see and re-plot.
```{r}
attenu %>% ggplot(aes(x = dist, y = accel)) + geom_point(aes(col = event)) + scale_x_log10() + scale_y_log10()
```

b) Fit a mixed effects model with the transformed variables, taking into account event and stations as random effects. Express the effect of magnitude on the acceleration. 
```{r}
model2 = lmer(log(accel) ~ log(dist) + mag + (1|event)+(1|station), data = attenu)
summary(model2)
```

c) Does adding a quadratic term in distance improve the model?
> adding a quadratic term does improve the model. 

```{r}
model3 = lmer(log(accel) ~ log(dist)+ I((log(dist)^2))+ mag + (1|event)+(1|station), data = attenu)
summary(model3)

fitted2 = predict(model2)
attenu %>% summarize(MSE = mean((fitted2-log(accel))^2), MAE = mean(abs(fitted2-log(accel))))

fitted3 = predict(model3)
attenu %>% summarize(MSE = mean((fitted3-log(accel))^2), MAE = mean(abs(fitted3-log(accel))))
```

d) Can we remove the station variation term? Explain why or why not.

```{r}
model_alt2 = lmer(log(accel) ~ log(dist)+ I((log(dist)^2))+ mag + (1|event)+(1|station), data = attenu, REML = FALSE)
model_null2 = model3 = lmer(log(accel) ~ log(dist)+ I((log(dist)^2))+ mag + (1|event), data = attenu, REML = FALSE)

LRT = as.numeric(2 *(logLik(model_alt2) - logLik(model_null2)))
1 - pchisq(LRT, df = 1)
```

e) For a new magnitude 6 earthquake, predict the acceleration for up to a distance of 200 miles. Make a plot of the data and add your predicted curve on top in a different color.

```{r}
newdata = expand.grid(mag = 6, dist = 1:200)
head(newdata)

prediction = predict(model2, newdata, re.form=~0)
p = ggplot() + geom_point(data = attenu, aes(x = log(dist), y = log(accel)), col = "blue") + geom_line(data = newdata, aes(x = log(dist), y = prediction), col = "red")

print(p)
```

f) Predict how the acceleration varied for the first event, where only one observation was available. Show the predicted acceleration up to 200 miles in a plot.

```{r}
newdata2 = expand.grid(mag = 7, dist = 1:200, event = 1, station = 117)
head(newdata2)

prediction = predict(model2, newdata2)

p2 = ggplot() + geom_point(data = attenu, aes(x = log(dist), y = log(accel)), col = "blue") + geom_line(data = newdata2, aes(x = log(dist), y = prediction), col = "red")

print(p2)
```

g) Compare your prediction plots. What about the plots is "similar", and what is "different"?

> Between the two plots, they are similar in terms of the trend of the straight line. They differ in that the scatter plot in the first plot has a tighter fit where as the second model has a losser fit (more outliers). Overall, the second prediction plot is more accrate than the first based on the fit of the line. 