---
title: "Lab 1"
author: "Emmanuel Rayappa"
date: "Updated `r Sys.Date()`"
output:
  pdf_document: null
  html_document:
    df_print: paged
subtitle: 'MTH 362: Statistical Modeling'
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(palmerpenguins)
library(fivethirtyeight)
library(tidyverse)
data(penguins)
data(spi_global_rankings)
```

Answer the questions below (use blockquotes to denote your answer by putting a ">" at the beginning of the line). When you're finished:

1.  Change your name in the "author:" space at the top of this document.
2.  Click the "Knit" button at the top to create an PDF/html file with your answers.
3.  Review your answers, make any changes as necessary, and re-"Knit".
4.  Save your file and upload to BlueLine.

# Interactions

When we introduced the `penguins` data set in class, we saw that a linear regression model worked well to describe the relationship between bill length and flipper length for three species of penguins.

```{r}
penguins %>%
  ggplot(aes(x=bill_length_mm, 
             y=flipper_length_mm)) + 
  geom_point() + 
  geom_smooth(method='lm',
              se=FALSE) +
  labs(x='Bill length (mm)', 
       y='Flipper length (mm)', 
       title='Palmer penguins')
```

Does the "best" linear model depend on which penguin species?

```{r}
penguins %>%
  ggplot(aes(x=bill_length_mm, 
             y=flipper_length_mm)) + 
  geom_point(aes(col=species, pch=species)) + 
  geom_smooth(method='lm',
              se=FALSE) +
  labs(x='Bill length (mm)', 
       y='Flipper length (mm)', 
       title='Palmer penguins')
```

---

There are two options we could pursue:

1. Fit a linear regression model on each species separately.
2. Use an __interaction term__ to add species to a multiple linear regression model.

__Interaction__: An interaction occurs when an explanatory variable has a different effect on the outcome depending on the values of another explanatory variable.

Does the relation between the explanatory and response variable depending on the `species`? 
> Write your response here.
Mybe, since the regression lines are not exactly the same.

```{r}
penguins %>%
  ggplot(aes(x=bill_length_mm, 
             y=flipper_length_mm, group = species)) + 
  geom_point(aes(col=species, pch=species)) + 
  geom_smooth(method='lm', se=FALSE) +
  labs(x='Bill length (mm)', 
       y='Flipper length (mm)', 
       title='Palmer penguins')
```

The linear model with interaction looks something like this:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_{12}(X_1 \times X_2) + \epsilon$$

In R, we specify an interaction term with an asterisk (`*`) or a colon (`:`).

The code below fits the linear regression model with an additional interaction term for `species`.

```{r}
model_int <- lm(flipper_length_mm ~ bill_length_mm*species, data=penguins)
summary(model_int)
```

a) Based on the output, write the linear model with the necessary dummy variables.

> Write your response here.

```{r}
model_out <- lm(flipper_length_mm ~ species, data=penguins)
summary(model_int)
```


b) Which terms in the model are "statistically significant"?

> Write your response here.
Bill length and species.

c) Evaluate the model assumptions.

> Write your response here.
Based on the residual plot, even though the independent assumption is not validated, the variance is not constant. Also, there are points that are based from the dash line in the qq plot, which means the normality assumption is not satisfied. Therefore, the linear model's assumption  is not reliable. 

```{r}
par(mfrow=c(2, 2))
plot(model_int)
```

d) Interpret the result

> Write your response here.
The model suggests that the bill length is positively related to the flipper length which means that 


# Club soccer predictions

FiveThirtyEight first started publishing club soccer predictions in January 2017. Currently, the model includes forecasts for club soccer teams in 36 leagues, as well as a strength measure called the "Soccer Power Index".

The R package `fivethirtyeight` contains the current SPI ratings and rankings for men's club teams (package last updated July 31, 2019).

FiveThirtyEight's website includes a description of what SPI is meant to represent:

*At the heart of our club soccer forecasts are FiveThirtyEight's SPI ratings, which are our best estimate of a team's overall strength. In our system, every team has an offensive rating that represents the number of goals it would be expected to score against an average team on a neutral field, and a defensive rating that represents the number of goals it would be expected to concede. These ratings, in turn, produce an overall SPI rating, which represents the percentage of available points --- a win is worth 3 points, a tie worth 1 point, and a loss worth 0 points --- the team would be expected to take if that match were played over and over again.*

The exact SPI formula is not published, but given the data from the `fivethirtyeight` package we can use linear models to "reverse engineer" it.

```{r}
glimpse(spi_global_rankings)
```

The package documentation includes the following description of each variable:

+-------------+-------------------------------------------------------------------------------------------+
| Variable    | Description                                                                               |
+=============+===========================================================================================+
| `name`      | The name of the soccer club.                                                              |
+-------------+-------------------------------------------------------------------------------------------+
| `league`    | The name of the league to which the club belongs.                                         |
+-------------+-------------------------------------------------------------------------------------------+
| `rank`      | A club's current global ranking.                                                          |
+-------------+-------------------------------------------------------------------------------------------+
| `prev_rank` | A club's previous global ranking.                                                         |
+-------------+-------------------------------------------------------------------------------------------+
| `off`       | Offensive rating for a given team (the higher the value the stronger the team's offense). |
+-------------+-------------------------------------------------------------------------------------------+
| `def`       | Defensive rating for a given team (the lower the value the stronger the team's defense).  |
+-------------+-------------------------------------------------------------------------------------------+
| `spi`       | A club's SPI score.                                                                       |
+-------------+-------------------------------------------------------------------------------------------+

## Question 1

The code below creates a scatterplot showing the relationship between SPI score and offensive rating.

a)  Is there a relationship between SPI and offensive rating? If so, how would you describe it?

> Write your answer here.
There is a strong positive linear reltiaonship between SPI and offensive rating. 

```{r}
spi_global_rankings %>% ggplot(aes(x=off, y=spi)) + 
  geom_point() + 
  labs(x='Offensive Rating', y='SPI Score')
```

b)  Does this relationship depend on the league in which a team plays? Can you tell which leagues are "stronger" than others based on this plot?

> Write your answer here.
Being in different leauges does not necessairly make the strength of the relationship change. 

```{r}
spi_global_rankings %>% ggplot(aes(x=off, y=spi)) + 
  geom_point() + 
  facet_wrap(~league) + 
  labs(x='Offensive Rating', y='SPI Score')
```

c)  Use the `lm()` function to fit a linear model to predict SPI based on a team's offensive rating alone. Write the fitted model and interpret each of the regression coefficients.

> Write your answer here.
> spi = -3.2799 + 36.3314 * off
> The estimated $\beta_0$ is -3.2799, which means when the offensive rating is 0, the SPI should be -3.2799. The estimated $\beta_1$ is 36.3314, which means that when the offensive rating increases 1 unit, the SPI will increase 36.3314 units.

```{r}
model1 = lm(spi~off, data = spi_global_rankings)
summary(model1)
```


d)  How much variability in a team's SPI score can be explained using a team's offensive rating alone?

> Write your answer here.
91.55% of variability can be explained.

e)  Is this linear regression model "statistically significant"? Explain how you know.

> Write your answer here.
Yes. Since the only variable offensive rating has a small p-value, which indicates it will significantly affect SPI. 

f)  Evaluate the LINE assumptions. What potential issues exist with this linear model?

> Write your answer here.
There are some outliers in the data based on the residual plot. The normal Q-Q plot suggests this as well, which may not necessairly be true and could make our linear model look less accurate.

```{r}
par(mfrow = c(2,2))
plot(model1)
```


## Question 2 (complete by yourself)

Repeat Question 1, with a team's defensive rating as the lone explanatory variable.

a)  Is there a relationship between SPI and offensive rating? If so, how would you describe it?
> Teams with highler defensive ratings tend to have lower SPI's. There is a Strong Linear Relationship.

```{r}
spi_global_rankings %>% ggplot(aes(x=def, y=spi)) + 
  geom_point() + 
  labs(x='Defensive Rating', y='SPI Score')
```
b)  Does this relationship depend on the league in which a team plays? Can you tell which leagues are "stronger" than others based on this plot?

> Write your answer here.
Being in different leauges does not cause the strength of the relationship change. 

```{r}
spi_global_rankings %>% ggplot(aes(x=def, y=spi)) + 
  geom_point() + 
  facet_wrap(~league) + 
  labs(x='Defensive Rating', y='SPI Score')
```
c)  Use the `lm()` function to fit a linear model to predict SPI based on a team's offensive rating alone. Write the fitted model and interpret each of the regression coefficients.

> Write your answer here.
> spi = 97.6605 + -38.8943 * def
> The estimated $\beta_0$ is 97.6605, which means when the defensive rating is 0, the SPI should be -38.8943. The estimated $\beta_1$ is -38.8943, which means that when the defensive rating increases 1 unit, the SPI will decrease by -38.8943 units.

```{r}
model1_2 = lm(spi~def, data = spi_global_rankings)
summary(model1_2)
```
d)  How much variability in a team's SPI score can be explained using a team's defensive rating alone?
> We can explain about 87.57% of the variability.

e)  Is this linear regression model "statistically significant"? Explain how you know.
> This model is statisitically significant because the p-values are each less than 5%, which indicates that it is signifcant. 

f)  Evaluate the LINE assumptions. What potential issues exist with this linear model?
The outliers in the line indicate that normality does not exist.

```{r}
par(mfrow = c(2,2))
plot(model1_2)
```


par(mfrow = c(2,2))
plot(model1)


## Question 3

What happens when both offensive rating and defensive rating are used as explanatory variables?

a)  Is there a relationship between defensive rating and offensive rating? If so, how would you describe it?

> Write your answer here.
There does appear to be a relationship between offensive and defensive rating. The relationship is negative and the strength of the relationship appears to be on the stronger side. 

```{r}
spi_global_rankings %>% ggplot(aes(y=off, x=def)) + 
  geom_point() + 
  labs(y='Offensive Rating', x='Defensive rating')
```


b)  Use the `lm()` function to fit a multiple linear model to predict SPI based on both team's offensive rating and defensive ratings. Write the fitted model and interpret each of the regression coefficients.

> Write your answer here.
spi = 42.2472 + 21.9535 * off - 19.3674 * def
The estimated $\beta_0$ is 42.2472, which means when the offensive rating is 0, the SPI should be 42.2472. The estimated $\beta_1$ is 21.9525, which means that when the the offensive rating increases 1 unit, the SPI will increase 21.9535 units. For the defensive rating, the $\beta_2$ is -19.3674, which means when the defnesive rating increases for one unit while the offsive rating stays the same, the SPI will decrease 19.3674 units.

```{r}
model2 = lm(spi~off + def, data = spi_global_rankings)
summary(model2)
```


c)  How much variability in a team's SPI score can be explained using the multiple linear regression model?

> Write your answer here.
98.95% can be explained using the linear regression model.

d)  Is this multiple linear regression model "statistically significant"? Explain how you know.

> Write your answer here.
Yes. Both variable have a small p-value, which means that they are statistically significant and will affect the SPI

e)  Evaluate the LINE assumptions. What potential issues exist with the multiple linear regression model?

> Write your answer here.
On the residual plot, there is a clear trend which means that we may miss some polynomial terms in the model. There are quite a few points which are off the dash line in the qq plot, which means that the normality assumption isn't satisfied.

```{r}
par(mfrow = c(2,2))
plot(model2)
```

## Question 4

One way to assess the fit of a multiple regression model is to plot the predicted values against the actual values, and look for points with high residuals.

a)  In the R code chunk below, replace the blanks with the name of your multiple regression model. Remove the `eval=FALSE` chunk option to run the code, and based on the resulting output, answer the question in the comment.

```{r}
names(model1)

# What does this line do?
spi_global_rankings <- spi_global_rankings %>%
  mutate(residuals= model1$residuals,
         fitted.values=model1$fitted.values)

glimpse(spi_global_rankings)
```

b)  Create a scatterplot of the predicted values from the multiple regression model against the actual values. Change the point color based on the magnitude of the residuals (*Hint: `abs()` is the absolute value function*). Comment on what you see.

> Write your answer here.
The plot does not do a good job 

```{r}
spi_global_rankings %>% ggplot(aes(y = fitted.values, x = spi)) + geom_point(aes(col = abs(residuals)))
```


c)  Which team(s) have the largest residuals? Explain why these teams might be poorly fit by the model.

> Write your answer here.
The teams with the highest net rating suprisingly have the lowest residuals, when we would expect this to be the opposite. 

```{r}
spi_global_rankings %>% arrange(residuals)
```
