---
title: "Lab 9"
author: "Emmanuel Rayappa"
date: "Updated `r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    theme: cosmo
    code_download: yes
subtitle: 'MTH 362: Statistical Modeling'
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

Answer the questions below (use blockquotes to denote your answer by putting a ">" at the beginning of the line). When you're finished:

1.  Change your name in the "author:" space at the top of this document.
2.  Click the "Knit" button at the top to create an PDF/HTML file with your answers.
3.  Review your answers, make any changes as necessary, and re-"Knit".
4.  Save your PDF/HTML file and upload to BlueLine.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(MASS)
library(faraway)
```

## Question 1

The `coagulation` data set comes from a study of blood coagulation times. Twenty-four animals were randomly assigned to four different diets, and the samples were taken in a random order.

```{r}
data(coagulation)
head(coagulation)
```

a) Make an appropriate plot of the data. Comment on what you see.
> A,B,and D don't have much differnece. C however is vastly different from the other 3 diets. 

```{r}
coagulation %>% ggplot(aes(x = diet, y = coag, group = diet)) + geom_boxplot(aes(col = diet))
```

b) Fit a fixed effects model and construct a prediction and 95% prediction interval for the response of a new animal assigned to diet D.
```{r}
model1 = lm(coag~diet, data = coagulation)
summary(model1)
predict(model1, newdata = data.frame(diet = "D"), interval = "prediction", levels = 0.95)
```

c) Fit a random effects model using REML. A new animal is assigned to diet D. Predict the blood coagulation time for this animal along with a 95% prediction interval.
```{r}
library(lme4)
model2=lmer(coag~1+(1|diet), data = coagulation)
summary(model2)
library(merTools)
predict(model2,newdata = data.frame(diet = "D"),interval = "prediction", levels = 0.95)
```

d) A new diet is given to a new animal. Predict the blood coagulation time for this animal along with a 95% prediction interval.

```{r}
predict(model2, newdata = data.frame(diet = "E"), allow.new.levels = TRUE)

model3 = lm(coag~1, data = coagulation)
predict(model3,interval = "prediction", levels = 0.95)
```

e) Which model is more appropriate: the fixed effects or random effects model? Explain your reasoning.
> It depends on whether or not we want to compare the 4 diets or we want to consider the possibility of another diet.If we only wnat to comapre the 4 diets, use the fixed effect. Otherwise,huse Random effect. 

## Question 2

The `babyfood` data set comes from a study on infant respiratory disease, namely the proportions of children developing bronchitis or pneumonia in their first year of life by type of feeding and sex.

```{r}
data(babyfood)
babyfood
```

a) Plot the proportion of infants with disease and gender. Comment what you have seen. 
```{r}
babyfood = babyfood %>% mutate(prop = (disease/(disease + nondisease)))
babyfood %>% ggplot(aes(x = sex, y = prop)) + geom_boxplot()
```

b) Compute the mean proportion of infants with disease with different food choices.
```{r}
babyfood %>% group_by(food)%>%summarise(mean = mean(prop))
```

c) Fit an appropriate mixed effects model by treating the food as a random effect. 
```{r}
model4 = glmer(prop~sex +(1|food), data = babyfood, family=binomial, weights = (disease + nondisease))
summary(model4)
```

d) Test whether there may be no variation attributable to sex.
> There is variation attributable to sex as the p-value is less than 0.05.

e) Test whether there may be no variation attributable to food.
```{r}
model5 = glm(prop~sex,data = babyfood, family=binomial, weights = (disease + nondisease))
summary(model5)
LRT = as.numeric(2 *(logLik(model4) -logLik(model5)))
1-pchisq(LRT, df = 1)
```


## Question 3 (complete by yourself)

A number of growers supply broccoli to a food processing plant. The plant instructs the growers to pack the broccoli into standard-size boxes. There should be 18 clusters of broccoli per box. Because the growers use different varieties and methods of cultivation, there is some variation in the cluster weights. The plant manager selected three growers at random and then four boxes at random supplied by these growers. Three clusters were selected from each box. The data may be found in the broccoli data set. The weight in grams of the cluster is given.

```{r}
data(broccoli)
head(broccoli)
```

a) Plot the data and comment on the nature of the variation seen.
> Grower 1 has the highest average weight while grower 3 has the second higest and grower 2 has the lowest average weight. 

```{r}
broccoli %>% ggplot(aes(x = grower, y = wt)) + geom_boxplot()
```

b) Compute the mean weights within growers. Compute the mean weights within boxes.
```{r}
broccoli %>% group_by(grower)%>%summarise(mean = mean(wt))
broccoli %>% group_by(box)%>%summarise(mean = mean(wt))
```

c) Fit an appropriate mixed effects model. Comment on how the variation is assigned to the possible sources.
> The variation is assigned based on the box number.

```{r, message=FALSE, warning=FALSE}
model6 = lmer(wt~1 + (1|grower)+(1|grower:box)+(1|grower:cluster),data = broccoli)
summary(model6)
```

d) Test whether there may be no variation attributable to growers.

```{r}
broccoli %>% ggplot(aes(x = box, y = wt)) + geom_point(aes(col = grower))
```

> There is no variation attribuatle to growers as the p-value is larger than 0.05 (in fact, it's at exactly 1)

e) Test whether there may be no variation attributable to boxes.
> There is no variation attributable to boxes. 

```{r}
model7 = lmer(wt~1 +(1|cluster)+(1|box), data = broccoli, REML = FALSE)
summary(model7)
LRT = as.numeric((2 *(logLik(model6))-(logLik(model7))))
1-pchisq(LRT, df = 1)
```


## Question 4

Last week we have identified the random effect and fixed effect through a couple of cases. This week we will try to write the _functional form_ of the models in R.

### Scenario A

Ten clinics are randomly sampled from a population. At each clinic, volunteers are randomly assigned into two groups. At each clinic, volunteers are randomly assigned into two groups. One group receives treatment 1, and the other group receives treatment 2. Let $N_{ij}$ be the number of patients receiving the $i^{th}$ treatment at the $j^{th}$ clinic. The response variable is $Y_{ij}$, the number of patients showing a favorable outcome to the treatment.

```{r, eval=FALSE}
treatment: fixed
clinic: random

glmer((Y_ij/N_ij) ~ treatment + (1|clinic), data = data, family = binomial, weight = N_ij)
```


### Scenario B

Seven locations are sampled. Each location is divided into four parts. Each part is randomly assigned a "level of exposure" to an environmental hazard (the levels are 0, 0.5X, X, and 2X), where X is the "nominal maximum safe level of exposure"). The response variable is the number of living organisms of a species affected by the hazard. The theory to be tested is that the number of organisms decreases linearly as the level of exposure increases.
```{r, eval=FALSE}
location: random
parts: nested with location
level of exposure: fixed
number or organisms: output

lmer(organisms~level + (1|location) + (1|location:parts), data = data, family = poisson)
  
```




### Scenario C

The country of Statsylvania has three distinct soil types. One location of each soil type is selected. Each location is split into 10 parts, called "plots". Five are randomly assigned to treatment 1, the other five to treatment 2. Corn is grown on each plot: the response of interest is the corn yield suitable for ethanol production: a total yield measurement is obtained for each plot. The objective is to estimate the affect of soil type and treatment on yield.

```{r,eval=FALSE}
loction: random
parts: nested with location
treatment: fixed
yield: output

lmer(yield ~ treatment + (1|location)+(1|location + parts), data = data)
```



### Scenario D

A court wants to determine if there is evidence of racial discrimination in the state's sentencing practices for murder convictions. Records for a period the court deems relevant are assembled and reviewed for race of the victim (white v. non-white), race of the convict (white v. non-white), and sentence type (death penalty v. no death penalty).
```{r, eval=FALSE}
victim:fixed
convited: fixed
period:random
type:output

glm(type ~ victim + convict + convict:victim + (1|period), data = data, family = binomial)
```



### Scenario E

Eighteen counties are randomly selected from a population of "typical" counties in america. Two types of vaccines will be compared for their effectiveness against the flu. Nine counties are assigned to vaccine type A, and nine are assigned to vaccine type B. In each county, the patients are randomly assigned to three groups. The first group receives the assigned vaccine at a low dose, the second group receives the vaccine at a medium dose, and the third group receives the assigned vaccine at a high dose. Let $N_{ijk}$ represent the number of patients in county $k$ receiving the $j^{th}$ dose level (low, medium, or high) of the $i^{th}$ vaccine type (A or B). Let $Y_{ijk}$ be the number of patients in the ${ijk}^{th}$ group who contract the flu. The goal of the study is to assess the effect of vaccine type and dose level on the effectiveness of protection.
```{r, eval=FALSE}
conties: random
type: fixed
group: fixed
output: proportion

glmer((y_ij/N_ij) ~ type + group + type:group + (1|countries), data=data, family = binomial, weights = N_ij)
```



### Scenario F

Twelve schools are sampled, one per county, from the same population in Scenario D. Six schools participate in a professional development program for math teachers; six do not (schools are randomly assigned to the development or no-development group). Students in the schools' math courses are given math proficiency tests at four times during the school year (at the beginning, after the first term, after the second term, and at the end of the school year). The objective is to see if the professional development program improved student learning in math.
```{r, eval=FALSE}
score: response
programs : fixed
school = random
lmer(score~program ~ (1|school), data = data)
```

